---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# epiUtils

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/rmgpanw/epiUtils/graph/badge.svg)](https://app.codecov.io/gh/rmgpanw/epiUtils)
[![R-CMD-check](https://github.com/rmgpanw/epiUtils/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/rmgpanw/epiUtils/actions/workflows/R-CMD-check.yaml)
[![pkgdown.yaml](https://github.com/rmgpanw/epiUtils/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/rmgpanw/epiUtils/actions/workflows/pkgdown.yaml)
<!-- badges: end -->

`epiUtils` provides essential tools for epidemiological analysis, including age-standardised rate calculations and person-years computation for cohort studies. The package implements established epidemiological methods with modern R practices, comprehensive validation, and built-in standard population data.

## Key Features

- **Age-Standardised Rates**: Calculate directly standardised rates with gamma or Byar's confidence intervals
- **Person-Years Calculation**: Compute person-years by age strata with automatic age-transition handling
- **WHO Standard Population**: Built-in WHO 2000-2025 World Standard Population data

## Installation

You can install the development version of epiUtils from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("rmgpanw/epiUtils")
```

## Quick Start

### Age-Standardised Rates

Calculate age-standardised incidence rates using the direct standardisation method:

```{r asr-example}
library(epiUtils)

# Example cancer incidence data
cancer_data <- data.frame(
  age_group = c("0-19", "20-39", "40-59", "60-79", "80+"),
  events = c(2L, 8L, 45L, 120L, 89L),
  person_years = c(50000, 65000, 55000, 40000, 15000),
  standard_pop = c(35000, 30000, 25000, 20000, 10000)
)

# Calculate age-standardised rates
result <- calculate_asr_direct(cancer_data)

# View results
cat("Age-Standardised Rate:", round(result$asr_scaled, 1), "per 100,000\n")
cat("95% CI:", round(result$ci_lower_scaled, 1), "-", round(result$ci_upper_scaled, 1), "\n")
```

### Person-Years Calculation

Calculate person-years of follow-up by age strata for cohort studies:

```{r person-years-example}
# Example cohort data
cohort_data <- data.frame(
  patient_id = 1:3,
  dob = as.Date(c("1960-01-15", "1955-06-20", "1972-09-01")),
  study_entry_date = as.Date(c("2020-03-01", "2021-01-01", "2019-01-01")),
  study_exit_date = as.Date(c("2024-06-15", "2025-12-31", "2022-04-10")),
  event_status = c(1L, 0L, 1L)
)

# Calculate person-years by age groups
py_result <- calculate_person_years_by_age_strata(cohort_data)
print(py_result)
```

### WHO Standard Population

Access the built-in [WHO 2000-2025 World Standard Population](https://seer.cancer.gov/stdpopulations/world.who.html):

```{r who-data-example}
# Load WHO standard population
data("who_2000_2025_standard_population")

# View structure
head(who_2000_2025_standard_population)
```

## Advanced Features

### Smart Case Count Handling

Warnings are raised for small case counts:

```{r warnings-example}
# Data with small case counts
small_counts_data <- data.frame(
  age_group = c("0-20", "20-40", "40-60"),
  events = c(0L, 3L, 15L),  # Zero and small case counts
  person_years = c(10000, 15000, 20000),
  standard_pop = c(15000, 20000, 25000)
)

# Calculate with warnings (default)
result_with_warnings <- calculate_asr_direct(small_counts_data)

# Calculate silently when appropriate
result_silent <- calculate_asr_direct(small_counts_data, warn_small_cases = FALSE)
```

### Confidence Interval Methods

Choose between two confidence interval calculation methods:

```{r ci-methods-example}
# Gamma method (default) - more conservative for small counts and low rates
result_gamma <- calculate_asr_direct(small_counts_data, warn_small_cases = FALSE)

# Byar's/Dobson method - traditional epidemiological approach  
result_byars <- calculate_asr_direct(small_counts_data, ci_method = "byars", warn_small_cases = FALSE)

cat("Gamma CI:", round(result_gamma$ci_lower_scaled, 1), "-", round(result_gamma$ci_upper_scaled, 1), "\n")
cat("Byar's CI:", round(result_byars$ci_lower_scaled, 1), "-", round(result_byars$ci_upper_scaled, 1), "\n")
```

### Integration with Standard Populations

Easily combine your data with WHO standard populations:

```{r integration-example}
# Using WHO standard population directly
who_subset <- who_2000_2025_standard_population[1:10, ]

# Create matching incidence data
incidence_data <- who_subset |>
  dplyr::mutate(
    events = as.integer(c(5, 8, 12, 18, 25, 35, 45, 60, 75, 85)),
    person_years = c(50000, 55000, 60000, 58000, 52000, 
                     48000, 42000, 35000, 25000, 15000),
    standard_pop = standard_for_seer_stat
  ) |>
  dplyr::select(age_group, events, person_years, standard_pop)

# Calculate ASR
who_result <- calculate_asr_direct(incidence_data)
cat("ASR using WHO standard:", round(who_result$asr_scaled, 1), "per 100,000\n")
```

## Citation

When using `epiUtils` in publications, please cite:

```{r citation, eval=FALSE}
citation("epiUtils")
```

## Contributing

Contributions are welcome! Please see the contributing guidelines and submit issues or pull requests on GitHub.
