---
title: "Getting Started with epiUtils"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with epiUtils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(epiUtils)
library(dplyr)
library(ggplot2)
```

# Introduction

The `epiUtils` package provides essential tools for epidemiological analysis, including:

- Age-standardised rate calculations using direct standardisation
- Person-years computation for cohort studies with age stratification
- Built-in WHO 2000-2025 World Standard Population data

This vignette demonstrates the key functions using practical examples.

# WHO World Standard Population

The package includes the WHO 2000-2025 World Standard Population data, which is commonly used for age standardisation in epidemiological studies.

```{r who-data}
# Load the WHO standard population data
data("who_2000_2025_standard_population")

# View the structure
head(who_2000_2025_standard_population)

# Summary of the data
str(who_2000_2025_standard_population)
```

# Example 1: Calculate Person-Years by Age Strata

Let's start with an example of calculating person-years for a cohort study. The default age groups now match the WHO 2000-2025 Standard Population, making it easy to calculate age-standardised rates.

```{r person-years-example}
# Create example cohort data
set.seed(123)
n_patients <- 100

cohort_data <- data.frame(
  patient_id = 1:n_patients,
  dob = as.Date("1950-01-01") + sample(-7300:7300, n_patients, replace = TRUE), # Random dates ±20 years
  study_entry_date = as.Date("2020-01-01") + sample(0:365, n_patients, replace = TRUE),
  study_exit_date = as.Date("2023-01-01") + sample(0:730, n_patients, replace = TRUE),
  event_status = sample(c(0L, 1L), n_patients, replace = TRUE, prob = c(0.8, 0.2))
)

# Ensure study_exit_date is after study_entry_date
cohort_data$study_exit_date <- pmax(
  cohort_data$study_exit_date, 
  cohort_data$study_entry_date + 30
)

# View first few rows
head(cohort_data)

# Calculate person-years by 5-year age groups (default - matches WHO age groups)
person_years_who <- calculate_person_years_by_age_strata(cohort_data)
print("Person-years by WHO-compatible age groups:")
print(person_years_who)

# Calculate person-years using custom age groups (10-year groups for comparison)
person_years_10yr <- calculate_person_years_by_age_strata(
  cohort_data, 
  age_cut_points = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 150)
)
print("Person-years by 10-year age groups:")
print(person_years_10yr)
```

# Example 2: Age-Standardised Rates with WHO Standard Population

Now let's calculate age-standardised incidence rates using the WHO standard population. With the updated defaults, this is now much simpler since the age groups match automatically.

```{r asr-example}
# Method 1: Direct calculation using WHO-compatible age groups (recommended)
# The default age groups now match WHO 2000-2025 Standard Population exactly

# First, let's verify the age groups match
print("Age groups from our person-years calculation:")
print(levels(person_years_who$age_group))

print("WHO 2000-2025 Standard Population age groups:")
print(who_2000_2025_standard_population$age_group)

print("Age groups match perfectly:", 
      identical(levels(person_years_who$age_group), who_2000_2025_standard_population$age_group))

# Merge with WHO standard population data directly
asr_data_who <- person_years_who |>
  left_join(
    who_2000_2025_standard_population |>
      select(age_group, standard_pop = standard_for_seer_stat),
    by = "age_group"
  ) |>
  filter(person_years > 0) |>  # Ensure positive person-years for all groups
  select(age_group, events, person_years, standard_pop)

print("Data prepared for ASR calculation (direct WHO matching):")
print(asr_data_who)

# Calculate age-standardised rates
asr_result_who <- calculate_asr_direct(asr_data_who)

print("Age-Standardised Rate Results (WHO age groups):")
print(paste("Crude rate:", round(asr_result_who$crude_rate_scaled, 2), "per 100,000"))
print(paste("Age-standardised rate:", round(asr_result_who$asr_scaled, 2), "per 100,000"))
print(paste("95% CI:", round(asr_result_who$asr_ci_lower_scaled, 2), "-", round(asr_result_who$asr_ci_upper_scaled, 2)))
print(paste("Total events:", asr_result_who$total_events))
print(paste("Total person-years:", asr_result_who$total_person_years))

# Method 2: Using custom age groups (for comparison)
# If you need custom age groups, you can still aggregate WHO data to match

# Create a mapping to match our 10-year age groups to WHO standard population
who_aggregated <- who_2000_2025_standard_population |>
  mutate(
    age_group_10yr = case_when(
      lower_age_limit < 10 ~ "[0-10)",
      lower_age_limit < 20 ~ "[10-20)",
      lower_age_limit < 30 ~ "[20-30)",
      lower_age_limit < 40 ~ "[30-40)",
      lower_age_limit < 50 ~ "[40-50)",
      lower_age_limit < 60 ~ "[50-60)",
      lower_age_limit < 70 ~ "[60-70)",
      lower_age_limit < 80 ~ "[70-80)",
      lower_age_limit < 90 ~ "[80-90)",
      lower_age_limit >= 90 ~ "[90+)"
    )
  ) |>
  group_by(age_group_10yr) |>
  summarise(
    standard_pop = sum(standard_for_seer_stat),
    .groups = "drop"
  )

# Merge with our 10-year person-years data
asr_data_10yr <- person_years_10yr |>
  left_join(who_aggregated, by = c("age_group" = "age_group_10yr")) |>
  filter(!is.na(standard_pop)) |>  # Remove any age groups not in WHO data
  filter(person_years > 0) |>  # Ensure positive person-years for all groups
  select(age_group, events, person_years, standard_pop)

print("Data prepared for ASR calculation (10-year age groups):")
print(asr_data_10yr)

# Calculate age-standardised rates using 10-year age groups
asr_result_10yr <- calculate_asr_direct(asr_data_10yr)

# Display comparison between WHO and 10-year age group results
print("Age-Standardised Rate Results (10-year age groups):")
print(paste("Crude rate:", round(asr_result_10yr$crude_rate_scaled, 2), "per 100,000"))
print(paste("Age-standardised rate:", round(asr_result_10yr$asr_scaled, 2), "per 100,000"))
print(paste("95% CI:", round(asr_result_10yr$asr_ci_lower_scaled, 2), "-", round(asr_result_10yr$asr_ci_upper_scaled, 2)))

# Compare the two approaches
print("Comparison of age standardisation approaches:")
print(paste("WHO age groups ASR:", round(asr_result_who$asr_scaled, 2), "per 100,000"))
print(paste("10-year age groups ASR:", round(asr_result_10yr$asr_scaled, 2), "per 100,000"))

# View age-specific details from WHO age groups
print("Age-specific details (WHO age groups):")
print(head(asr_result_who$age_specific_data[[1]], 10))

# Note: The WHO age groups provide more precise age standardisation
# while custom age groups may be useful for specific analytical needs
```

## Handling Small Case Counts

The `calculate_asr_direct()` function includes built-in warnings for small case counts, which help users make informed decisions about age grouping strategies. The warnings are raised regardless of the confidence interval method used.

```{r small-cases-demo}
# Create example data with small case counts to demonstrate warnings
small_cases_data <- data.frame(
  age_group = c("0-20", "20-40", "40-60", "60-80"),
  events = c(0L, 2L, 4L, 15L),  # Mix of zero and small case counts
  person_years = c(10000, 15000, 20000, 18000),
  standard_pop = c(25000, 30000, 20000, 15000)
)

print("Data with small case counts:")
print(small_cases_data)

# Calculate ASR - will show warnings about small case counts
print("With warnings (default):")
result_with_warnings <- calculate_asr_direct(small_cases_data)

# Calculate ASR with warnings disabled
print("With warnings disabled:")
result_no_warnings <- calculate_asr_direct(small_cases_data, warn_small_cases = FALSE)

# Results are identical regardless of warning setting
print(paste("ASR:", round(result_with_warnings$asr_scaled, 2), "per 100,000"))
```

## Confidence Interval Methods

The function supports two methods for calculating confidence intervals, controlled by the `ci_method` parameter. Understanding the differences helps you choose the most appropriate method for your analysis.

### Method Overview

1. **"gamma"** (default): Uses gamma distribution approach
2. **"byars"**: Uses Byar's approximation with Dobson adjustment

### Technical Differences

**Gamma Method:**

- Based on the gamma distribution, which naturally models positive rate data
- Automatically ensures confidence intervals are always positive (cannot be negative)
- Particularly robust with small case counts and low rates
- Consistent with `epitools::ageadjust.direct()`
- Uses shape and scale parameters derived from the observed data

**Byar's/Dobson Method:**

- Traditional epidemiological approach dating to the 1970s-80s
- Uses normal approximation with continuity corrections for larger counts
- Uses exact Poisson-based intervals for very small counts (<10 total cases)
- Consistent with `PHEindicatormethods::calculate_dsr()`

```{r ci-methods-demo}
# Compare confidence interval methods using the same data
print("Gamma method (default):")
result_gamma <- calculate_asr_direct(small_cases_data, warn_small_cases = FALSE)
print(paste("ASR:", round(result_gamma$asr_scaled, 2), 
            "95% CI:", round(result_gamma$asr_ci_lower_scaled, 2), "-", 
            round(result_gamma$asr_ci_upper_scaled, 2)))

print("Byar's/Dobson method:")
result_byars <- calculate_asr_direct(small_cases_data, ci_method = "byars", warn_small_cases = FALSE)
print(paste("ASR:", round(result_byars$asr_scaled, 2), 
            "95% CI:", round(result_byars$asr_ci_lower_scaled, 2), "-", 
            round(result_byars$asr_ci_upper_scaled, 2)))

# The point estimates are identical, but confidence intervals may differ
print("Point estimates are identical:", 
      round(result_gamma$asr_scaled, 10) == round(result_byars$asr_scaled, 10))

# Calculate the difference in CI width
gamma_width <- result_gamma$asr_ci_upper_scaled - result_gamma$asr_ci_lower_scaled
byars_width <- result_byars$asr_ci_upper_scaled - result_byars$asr_ci_lower_scaled
print(paste("CI width difference:", round(abs(gamma_width - byars_width), 2)))
```

### Practical Implications

**When results differ:**

- Differences are typically small for moderate to large case counts (>20 total cases)
- More noticeable differences may occur with very small case counts or low rates
- The gamma method tends to be slightly more conservative (wider intervals) with extremely small counts
- Both methods are statistically valid and widely accepted

**Method selection guidance:**

- Use **gamma** (default) for most analyses, especially with small case counts or low rates
- Use **byars** when:
  
  - Comparing with historical studies that used Byar's method
  - Institutional requirements specify this traditional approach
  
- Both methods converge to similar results as sample sizes increase

# Example 3: Using WHO Standard Population Directly

For a more straightforward example, let's create some synthetic cancer incidence data that directly matches the WHO standard population age groups.

```{r direct-who-example}
# Create synthetic cancer incidence data using WHO age groups
set.seed(456)

# Use WHO age groups directly (first 15 age groups, up to 75+)
cancer_data <- who_2000_2025_standard_population |>
  slice(1:15) |>  # Use first 15 age groups (0-74)
  mutate(
    # Simulate population sizes (person-years)
    person_years = round(runif(n(), min = 50000, max = 200000)),
    # Simulate cancer cases with age-specific rates
    # Higher rates in older age groups
    age_specific_rate = case_when(
      lower_age_limit < 20 ~ runif(1, 0.00001, 0.00005),
      lower_age_limit < 40 ~ runif(1, 0.00005, 0.0002),
      lower_age_limit < 60 ~ runif(1, 0.0002, 0.001),
      TRUE ~ runif(1, 0.001, 0.003)
    ),
    events = as.integer(round(person_years * age_specific_rate))
  ) |>
  select(age_group, events, person_years, standard_pop = standard_for_seer_stat)

print("Synthetic cancer incidence data:")
print(cancer_data)

# Calculate age-standardised incidence rates
cancer_asr <- calculate_asr_direct(cancer_data)

print("Cancer Incidence Results:")
print(paste("Crude incidence rate:", round(cancer_asr$crude_rate_scaled, 1), "per 100,000"))
print(paste("Age-standardised incidence rate:", round(cancer_asr$asr_scaled, 1), "per 100,000"))
print(paste("95% CI:", round(cancer_asr$asr_ci_lower_scaled, 1), "-", round(cancer_asr$asr_ci_upper_scaled, 1), "per 100,000"))
```

# Visualization

Let's create some visualizations to better understand our results.

```{r visualization}
# Plot age-specific rates
age_specific_data <- cancer_asr$age_specific_data[[1]]

# Age-specific rates plot
p1 <- ggplot(age_specific_data, aes(x = age_group, y = age_specific_rate_scaled)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Age-Specific Cancer Incidence Rates",
    x = "Age Group",
    y = "Rate per 100,000"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p1)

# Contribution to ASR plot
p2 <- ggplot(age_specific_data, aes(x = age_group, y = contribution_to_asr * 100000)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Contribution of Each Age Group to Age-Standardised Rate",
    x = "Age Group",
    y = "Contribution per 100,000"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p2)

# Compare crude vs age-standardised rates
comparison_data <- data.frame(
  Rate_Type = c("Crude Rate", "Age-Standardised Rate"),
  Rate = c(cancer_asr$crude_rate_scaled, cancer_asr$asr_scaled),
  CI_Lower = c(NA, cancer_asr$asr_ci_lower_scaled),
  CI_Upper = c(NA, cancer_asr$asr_ci_upper_scaled)
)

p3 <- ggplot(comparison_data, aes(x = Rate_Type, y = Rate)) +
  geom_col(fill = c("orange", "purple"), alpha = 0.7, width = 0.5) +
  geom_errorbar(
    aes(ymin = CI_Lower, ymax = CI_Upper), 
    width = 0.2, 
    na.rm = TRUE
  ) +
  theme_minimal() +
  labs(
    title = "Crude vs Age-Standardised Cancer Incidence Rates",
    x = "Rate Type",
    y = "Rate per 100,000"
  )

print(p3)
```

# Example 4: Age-Standardised Prevalence Rates

The `calculate_asr_direct()` function can also be used to calculate age-standardised prevalence rates. For prevalence studies, the key differences are:

- **events**: Number of individuals with the condition at a point in time
- **person_years**: Total number of individuals examined/surveyed (use population counts, not person-years of follow-up)
- **multiplier**: Use 100 to express results as percentages (%) instead of rates per 100,000

```{r prevalence-example}
# Example: Age-standardised prevalence of diabetes in a population survey
set.seed(789)

# Create synthetic diabetes prevalence data
diabetes_data <- data.frame(
  age_group = c("18-29", "30-39", "40-49", "50-59", "60-69", "70+"),
  
  # Number of individuals with diabetes (events)
  events = c(12L, 45L, 98L, 156L, 203L, 167L),
  
  # Total number of individuals surveyed in each age group (person_years = N individuals)
  person_years = c(1200, 1500, 1800, 1400, 1100, 800),
  
  # WHO standard population weights for corresponding age groups
  standard_pop = c(2435, 2021, 1826, 1485, 1118, 763)  # Simplified WHO weights
)

print("Diabetes prevalence survey data:")
print(diabetes_data)

# Calculate age-standardised prevalence using multiplier = 100 for percentages
diabetes_prevalence <- calculate_asr_direct(
  .df = diabetes_data,
  multiplier = 100,  # Use 100 to get percentages instead of per 100,000
  warn_small_cases = FALSE
)

print("Diabetes Prevalence Results:")
print(paste("Crude prevalence:", round(diabetes_prevalence$crude_rate_scaled, 2), "%"))
print(paste("Age-standardised prevalence:", round(diabetes_prevalence$asr_scaled, 2), "%"))
print(paste("95% CI:", round(diabetes_prevalence$asr_ci_lower_scaled, 2), "% -", 
            round(diabetes_prevalence$asr_ci_upper_scaled, 2), "%"))

# View age-specific prevalence rates
age_specific_prev <- diabetes_prevalence$age_specific_data[[1]]
print("Age-specific prevalence rates:")
print(age_specific_prev[, c("age_group", "events", "person_years", "age_specific_rate_scaled")])

# Plot age-specific prevalence
library(ggplot2)
p_prev <- ggplot(age_specific_prev, aes(x = age_group, y = age_specific_rate_scaled)) +
  geom_col(fill = "darkred", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Age-Specific Diabetes Prevalence",
    x = "Age Group",
    y = "Prevalence (%)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_prev)
```

**Key Points for Prevalence Studies:**

- **Cross-sectional data**: Use counts of individuals with/without condition at a specific time point
- **Population denominator**: Use total number of individuals examined, not person-years of follow-up
- **Percentage scale**: Set `multiplier = 100` to express results as percentages (%)
- **Interpretation**: Results represent the proportion of the population with the condition, standardised for age differences

# Key Features Summary

## `calculate_person_years_by_age_strata()`

- Calculates person-years of follow-up by age groups
- **Default age groups now match WHO 2000-2025 Standard Population exactly**
- Handles age transitions during follow-up period automatically
- Flexible age group definitions for custom analyses
- Final age group displays as "[100+)" to match WHO format
- Returns events and person-years by age strata

## `calculate_asr_direct()`

- Calculates age-standardised rates using direct standardisation
- **Works seamlessly with WHO age groups from person-years function**
- Supports both incidence rates and prevalence calculations
- Two confidence interval methods: gamma distribution (default) and Byar's/Dobson approximation
- Smart warnings for small/zero case counts with user control
- Returns comprehensive results including crude rates, ASRs, confidence intervals, and totals

## WHO Standard Population Data

- Ready-to-use WHO 2000-2025 World Standard Population
- Multiple standard population formats available
- Easily integrates with the package functions

This vignette demonstrates the basic usage of the key functions in `epiUtils`. For more detailed information about specific functions, see their individual help pages.
